Effects of the body force on the pedestrian and the evacuation dynamics

### 2020-03-04 paper.tex
quite el texto rojo y azul en todo el paper

The maximum (realistic) compressions may be computed from the Hooke's relation 
$F(x)=k_n(x)\,x$ and the compressibility $k_n(x)$ reported in Table~\ref{table_data}. 
An ``uncomfortable'' body force  $10\,$N $-$ $30\,$N can address compressions in the 
range of $0.030-0.055\,$m. Also, a ``hitting'' force of $60\,$N can address 
compressions between $0.045$ and $0.065\,$m. 
Besides, no reliable values for the sliding friction $k_t$ appears to be 
available in the literature (to our knowledge).\\
%Thus, according to Table~\ref{table_data}, \textcolor{blue}{we may expect experimental values for $k_n(x)$ up to 
%$1000\,$kg/s$^2$ (for $F=30\,$N), or, up to $1400\,$kg/s$^2$ (for $F=60\,$N).} \\

@@

We present in this section the results corresponding to the corridor geometry. 
We show the effects of \textcolor{blue}{\sout{modifying} varying} the body force coefficient $k_n$ on the 
collective dynamics. We will keep $v_d$ (or $\mathcal{A}$) 
fixed along this Section. \\

@@

We find these topological magnitudes useful for comparing the 
\textcolor{blue}{pedestrian behavior in the} 
corridor geometry with respect to the bottleneck geometry. A re-examination of 
Fig.~\ref{degree_dens} (corridor) and Fig.~\ref{degree_vd} (bottleneck) reveal 
that the pedestrian stiffness $k_n$ affects differently the way they contact 
each other. The mean degree increases for ``stiff'' pedestrians moving along 
the corridor \textcolor{blue}{\sout{(at densities above} 
as the density increased with a sharp increase at} $5\,$p/m$^2$\textcolor{blue}{\sout{)}}. Conversely, the mean degree 
increases for ``soft'' pedestrians in the bottleneck situation.\\

Fig.~\ref{network_bottleneck} and 
Fig.~\ref{network_corridor} illustrate the connectivity differences between 
the bottleneck and the corridor situation. The bulk in 
Fig.~\ref{network_bottleneck} appears more heavily connected among ``soft'' 
pedestrians than among ``stiff'' pedestrians (see both snapshots at $v_d=10$). 
The opposite occurs in Fig.~\ref{network_corridor}. This discrepancy seems to 
be related to the boundary conditions, since the same SFM parameters were 
applied on both situations. We may speculate that this phenomenon occurs 
because, in the corridor, the lateral walls act like a \textcolor{blue}{\sout{containment}
confining} barrier that 
forces the ``stiff'' pedestrian to increase his (her) contacts. On the contrary, 
no real containment walls exist in the bottleneck situation (regardless the side 
walls). \\

We can test our hypothesis by computing the mean overlap. 
Fig.~\ref{overlap_dens} shows this magnitude for the corridor situation. Notice 
that ``soft'' pedestrians attain \textcolor{blue}{\sout{a more intruding} more } overlap than the ``stiff'' 
ones, as expected. This is in agreement with Fig.~\ref{overlap_vd} for the 
bottleneck situation (at a fixed value of $v_d$). The curves in 
Fig.~\ref{overlap_vd}, however, do not meet each other as in  
Fig.~\ref{overlap_dens}. \textcolor{red}{\sout{The curve meeting attains for space limitations in 
the corridor, since overlapping occurs (almost) independently of the value of
$k_n$. This is clearly not the case in the bottleneck situation.} 
All the curves meet each other at high-density values. This phenomenon occurs
due to space limitations in the corridor that produces overlapping (almost) independently of the
value of $k_n$}. \\

@@

The above results \textcolor{blue}{\sout{correspond to a milestone} represent an important step} in the 
investigation. The \textcolor{blue}{inter-pedestrian} connectivity differences between the bottleneck geometry 
and the corridor geometry were not expected and opens two major questions:
\textcolor{blue}{\sout{ how 
does the body stiffness interact with the walls} how do the pedestrians interact with the walls as
a function of the body stiffness}, and consequently, how does 
this affect the flux across the corridor.\\

@@

The $\langle v_x\rangle$ values in Fig.~\ref{kn_vs_vx_corridor} 
were computed at densities $\rho\geq 5.5$. According to 
Fig.~\ref{overlap_dens},\textcolor{blue}{\sout{ these densities attain relevant overlaps}
at these densities, the overlaps among pedestrians attain significant values}. Furthermore, 
stiffness values $k_n\geq 10^4$ move the system to a more heavily connected 
stage, and thus, pedestrians have no choice but to walk at 
(almost) the same speed as his (her) neighbors. This 
behavior may be envisaged as the passage from a \textcolor{red}{\sout{released} ``free''} 
walking movement to a \textcolor{red}{\sout{tigh} constrained} movement as the stiffness increases. A more 
physical picture would assimilate te former as a ``fluid-like 
state'' and the latter as a ``solid-like state''. When the stiffness is 
very high (say $k_n=$1.2~E6), all the pedestrians 
are expected to walk at a common velocity. The 
pedestrians that walk in physical contact with the wall are the ones who 
determine the velocity of the whole crowd, as discussed 
below.\\

Fig.~\ref{vx_profile} shows the velocity profile ($\langle v_x \rangle$ vs. the transversal coordinate $y$) for different $k_n$ values. For $k_n=0$, we can see a parabolic-like velocity profile \textcolor{red}{which means that the friction with the walls reduces the speed of pedestrians.} \textcolor{blue}{\sout{that} The velocity profile} resembles the Poiseuille flow (similar to Newtonian and incompressible fluids in a laminar regime). This behavior was also observed in empirical measurements of crowd dynamics reported in Ref.~\cite{zhang2013empirical}. But as $k_n$ increases the velocity profile flattens until becoming (almost) uniform (see Fig.~\ref{vx_profile} for $k_n=1.2\,$E6). In this scenario, $v_x$ attains a much lower \textcolor{blue}{\sout{level} value}  than in the case of soft pedestrians ($k_n=$0).\\

\textcolor{blue}{From the results displayed above, we realize that \sout{The} the} crowd behaves like a solid for very stiff pedestrians. This means that the crowd can not be easily ``deformed''. In this context, deformation means that parts of the crowd may be \textcolor{blue}{\sout{released} allowed} to move faster than other parts of the crowd. \\

\textcolor{blue}{\sout{The strain rate is the tensor that expresses how the relative velocity of a medium changes when one moves away from one position to another.} The strain rate tensor displays the rate of change of the deformation of a body in the vincinity of a given point.} We consider the following discrete definition of the strain rate:

@@

We conclude this Section by stressing\textcolor{blue}{, once again,} that stiffer pedestrians attain opposite results in corridors with respect to bottlenecks. The walls in the corridors play a critical role that prevents pedestrians from detaching from each other. This effect has can be observed as an increment in the connectivity of the contact network and the flattening of the velocity profile (thus reducing the strain rate). Very high enough stiffness values stuck the pedestrians leading to a ``solidification'' of the crowd. All the pedestrians walk at almost the same velocity at this stage. The pedestrians that are in contact with the walls are the ones that determine the velocity of the whole crowd. We already analyzed this situation in Ref~\cite{dorso_2019}. \\

@@

\textcolor{blue}{Having adopted as fundamental philosophy of work that the parameters of the SFM should be properly determined by requiring that the model properly reproduces well established experimental data, we} explored the effect on the pedestrians dynamics of the (sometimes neglected) body force in the framework of the SFM. We showed that the stiffness coefficient ($k_n$) has a significant impact on the evacuation dynamics (bottleneck) and also in the dynamics of pedestrians walking along a straight corridor.\\

@@

The opposite behavior is obtained in the corridor geometry with respect to the bottleneck geometry. The major difference is that pedestrians are limited to the available space between walls in the corridor geometry.  Walls are not relevant in the bottleneck geometry. Thus, the overlap between pedestrians is controlled by the available space. But, stiffer pedestrians are more likely to get stuck. The whole crowd can be compared to a granular material. Granular materials can be disordered (amorphous) or ordered depending on how particles \textcolor{blue}{\sout{link} interact}  to each other. In the present context, at low stiffness levels, the crowd appears disordered attaining a parabolic velocity profile.
If the stiffness level is high, the whole crowd appears ordered into a lattice (like a crystalline solid) with a uniform velocity profile that depends on the friction interaction with the walls.\\

Our efforts to ``tune'' the original SFM to \textcolor{blue}{reproduce} empirical data (say, the fundamental diagram) moved us to explore the dimensionless parameter space.  We found that \textcolor{blue}{\sout{the qualitative behavior of empirical  data can be accomplished} we can qualitatively reproduce the empirical data} if the parameters are close to $\mathcal{K}_c < 68$, $\mathcal{K} = 685$ and $\mathcal{A}=14$. Nevertheless, this is a first attempt to arrive at suitable parameters, although we do not claim these to be optimal. We encourage further research to find dimensionless numbers that may better fit experimental data.\\

@@

The SFM description in section~\ref{sfm} introduces seven parameters ($m$, 
$\tau$, $v_d$, $B$, $A$, $k_t$ and $k_n$) attaining for the ``individual'' 
behavior of each pedestrian. The collective dynamic, however, requires a 
smaller set of parameters. In order to identify this smaller set, we introduce 
the following \textcolor{blue}{dimensionless} magnitudes\\

@@

The equation of motion (\ref{eqn_motion}) can be \textcolor{blue}{rewritten} in terms of these 
(\textcolor{blue}{dimensionless}) magnitudes, while only three (reduced) parameters are needed.\\

>>>>

The maximum (realistic) compressions may be computed from the Hooke's relation 
$F(x)=k_n(x)\,x$ and the compressibility $k_n(x)$ reported in Table~\ref{table_data}. 
An ``uncomfortable'' body force  $10\,$N $-$ $30\,$N can address compressions in the 
range of $0.030-0.055\,$m. Also, a ``hitting'' force of $60\,$N can address 
compressions between $0.045$ and $0.065\,$m. 
Besides, no reliable values for the sliding friction $k_t$ appears to be 
available in the literature (to our knowledge).\\

@@

We present in this section the results corresponding to the corridor geometry. 
We show the effects of varying the body force coefficient $k_n$ on the 
collective dynamics. We will keep $v_d$ (or $\mathcal{A}$) 
fixed along this Section. \\

@@

We find these topological magnitudes useful for comparing the 
pedestrian behavior in the 
corridor geometry with respect to the bottleneck geometry. A re-examination of 
Fig.~\ref{degree_dens} (corridor) and Fig.~\ref{degree_vd} (bottleneck) reveal 
that the pedestrian stiffness $k_n$ affects differently the way they contact 
each other. The mean degree increases for ``stiff'' pedestrians moving along 
the corridor as the density increased with a sharp increase at
 $5\,$p/m$^2$. Conversely, the mean degree 
increases for ``soft'' pedestrians in the bottleneck situation.\\

Fig.~\ref{network_bottleneck} and 
Fig.~\ref{network_corridor} illustrate the connectivity differences between 
the bottleneck and the corridor situation. The bulk in 
Fig.~\ref{network_bottleneck} appears more heavily connected among ``soft'' 
pedestrians than among ``stiff'' pedestrians (see both snapshots at $v_d=10$). 
The opposite occurs in Fig.~\ref{network_corridor}. This discrepancy seems to 
be related to the boundary conditions, since the same SFM parameters were 
applied on both situations. We may speculate that this phenomenon occurs 
because, in the corridor, the lateral walls act like a 
confining barrier that 
forces the ``stiff'' pedestrian to increase his (her) contacts. On the contrary, 
no real containment walls exist in the bottleneck situation (regardless the side 
walls). \\

We can test our hypothesis by computing the mean overlap. 
Fig.~\ref{overlap_dens} shows this magnitude for the corridor situation. Notice 
that ``soft'' pedestrians attain more overlap than the ``stiff'' 
ones, as expected. This is in agreement with Fig.~\ref{overlap_vd} for the 
bottleneck situation (at a fixed value of $v_d$). The curves in 
Fig.~\ref{overlap_vd}, however, do not meet each other as in  
Fig.~\ref{overlap_dens}. All the curves meet each other at
high-density values. This phenomenon occurs due to space limitations
in the corridor that produces overlapping (almost) independently of 
the value of $k_n$. \\

@@

The above results represent an important step in the 
investigation. The inter-pedestrian connectivity differences between the bottleneck geometry 
and the corridor geometry were not expected and opens two major questions:
 how do the pedestrians interact with the walls as
a function of the body stiffness, and consequently, how does 
this affect the flux across the corridor.\\

@@

The $\langle v_x\rangle$ values in Fig.~\ref{kn_vs_vx_corridor} 
were computed at densities $\rho\geq 5.5$. According to 
Fig.~\ref{overlap_dens},
at these densities, the overlaps among pedestrians attain significant values. Furthermore, 
stiffness values $k_n\geq 10^4$ move the system to a more heavily connected 
stage, and thus, pedestrians have no choice but to walk at 
(almost) the same speed as his (her) neighbors. This 
behavior may be envisaged as the passage from a ``free'' 
walking movement to a  constrained movement as the stiffness increases. A more 
physical picture would assimilate the former as a ``fluid-like 
state'' and the latter as a ``solid-like state''. When the stiffness is 
very high (say $k_n=$1.2~E6), all the pedestrians 
are expected to walk at a common velocity. The 
pedestrians that walk in physical contact with the wall are the ones who 
determine the velocity of the whole crowd, as discussed 
below.\\

Fig.~\ref{vx_profile} shows the velocity profile ($\langle v_x \rangle$ vs. the transversal coordinate $y$) for different $k_n$ values. For $k_n=0$, we can see a parabolic-like velocity profile which means that the friction with the walls reduces the speed of pedestrians. The velocity profile resembles the Poiseuille flow (similar to Newtonian and incompressible fluids in a laminar regime). This behavior was also observed in empirical measurements of crowd dynamics reported in Ref.~\cite{zhang2013empirical}. But as $k_n$ increases the velocity profile flattens until becoming (almost) uniform (see Fig.~\ref{vx_profile} for $k_n=1.2\,$E6). In this scenario, $v_x$ attains a much lower  value  than in the case of soft pedestrians ($k_n=$0).\\

From the results displayed above, we realize that the crowd behaves like a solid for very stiff pedestrians. This means that the crowd can not be easily ``deformed''. In this context, deformation means that parts of the crowd may be allowed to move faster than other parts of the crowd. \\

The strain rate tensor displays the rate of change of the deformation of a body in the vicinity of a given point.
 We consider the following discrete definition of the strain rate:

@@

We conclude this Section by stressing, once again, that stiffer pedestrians attain opposite results in corridors with respect to bottlenecks. The walls in the corridors play a critical role that prevents pedestrians from detaching from each other. This effect has can be observed as an increment in the connectivity of the contact network and the flattening of the velocity profile (thus reducing the strain rate). Very high enough stiffness values stuck the pedestrians leading to a ``solidification'' of the crowd. All the pedestrians walk at almost the same velocity at this stage. The pedestrians that are in contact with the walls are the ones that determine the velocity of the whole crowd. We already analyzed this situation in Ref~\cite{dorso_2019}. \\

@@

Having adopted as fundamental philosophy of work that the parameters of the SFM should be properly determined by requiring that the model properly reproduces well established experimental data, we explored the effect on the pedestrians dynamics of the (sometimes neglected) body force in the framework of the SFM. We showed that the stiffness coefficient ($k_n$) has a significant impact on the evacuation dynamics (bottleneck) and also in the dynamics of pedestrians walking along a straight corridor.\\

@@

The opposite behavior is obtained in the corridor geometry with respect to the bottleneck geometry. The major difference is that pedestrians are limited to the available space between walls in the corridor geometry.  Walls are not relevant in the bottleneck geometry. Thus, the overlap between pedestrians is controlled by the available space. But, stiffer pedestrians are more likely to get stuck. The whole crowd can be compared to a granular material. Granular materials can be disordered (amorphous) or ordered depending on how particles interact with each other. In the present context, at low stiffness levels, the crowd appears disordered attaining a parabolic velocity profile.
If the stiffness level is high, the whole crowd appears ordered into a lattice (like a crystalline solid) with a uniform velocity profile that depends on the friction interaction with the walls.\\

Our efforts to ``tune'' the original SFM to reproduce empirical data (say, the fundamental diagram) moved us to explore the dimensionless parameter space.  We found that  we can qualitatively reproduce the empirical data if the parameters are close to $\mathcal{K}_c < 68$, $\mathcal{K} = 685$ and $\mathcal{A}=14$. Nevertheless, this is a first attempt to arrive at suitable parameters, although we do not claim these to be optimal. We encourage further research to find dimensionless numbers that may better fit experimental data.\\

@@

The SFM description in section~\ref{sfm} introduces seven parameters ($m$, 
$\tau$, $v_d$, $B$, $A$, $k_t$ and $k_n$) attaining for the ``individual'' 
behavior of each pedestrian. The collective dynamic, however, requires a 
smaller set of parameters. In order to identify this smaller set, we introduce 
the following dimensionless magnitudes\\

@@

The equation of motion (\ref{eqn_motion}) can be rewritten in terms of these 
(dimensionless) magnitudes, while only three (reduced) parameters are needed.\\
